{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-01 thesis/tm.tex
buncha minor revisions tm

Support for TSX programs in Landslide is implemented in three parts, broadly speaking,
the user interface plus one internal part corresponding to each proof above.

@@

Landslide provides its own ``implementation'' of the TSX interface,
which matches GCC's interface exactly,
located in {\tt 410user/inc/htm.h}
under {\tt pebsim/p2-basecode/}.
The interface is implemented in {\tt 410user/libhtm/htm.c},
perhaps surprisingly, as totally empty functions:
Landslide hooks these functions' addresses during instrumentation
and inserts preemption points, failure injections,
et cetera as necessary whenever the execution encounters them.

@@

Finally, Quicksand supports the following command-line options to enable various sets of HTM features within Landslide.
Each option requires each of the previously-listed options to also be present.
\begin{itemize}
	\item {\tt -X} (for ``tsX'' or ``Xbegin'') enables the basic features:
		preemption points on each {\tt \_xbegin()} and {\tt \_xend()} call,
		failure injection on each of the former
		(always returning {\tt \_XABORT\_RETRY} as the failure code)
		(\sect{\ref{sec:txn-failure}}),
		and treatment of transactional regions during data-race analysis (\sect{\ref{sec:txn-datarace}}).
	\item {\tt -A} (for ``xAbort codes'')
		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
	\item {\tt -S} (for ``Stm'')
		disables the {\tt \_XABORT\_RETRY} failure reason,
		which causes Landslide to match STM's execution semantics rather than HTM.
		(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
\end{itemize}

@@

While Figure~\ref{fig:retry-set-example}'s example may seem contrived
(what program's transactions would just give up and do no work if aborted?),
it is easy to imagine a larger transactional data structure,
two insertions into which might operate on disjoint nodes or array indices,
allowing simultaneous transactions to usually succeed,
while the more uncommon abort paths might take the opportunity to assert a full consistency check of all elements,
ultimately resulting in a similar conflict pattern.
Also, the evaluation will later show
that even programs with fully conflicting success/abort paths
may still exhibit some equivalent thread interleavings of this pattern
after their failure paths are split apart into smaller transitions by data-race preemption points
(\sect{\ref{sec:tm-eval-retrysets}}).

To solve this,
I extended Landslide
with {\em retry set reduction},
an experimental feature named after the analogous {\em sleep sets} from prior work
(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
if either or both sides of a memory conflict were part of a transaction,
it records a {\em retry set},
i.e., the pair of {\tt xbegin} results executed by the conflicting threads.
% ah-- not compatible after all!
%in addition to tracking the set of abort codes possible at each preemption point.
Later, when traversing that new branch,
Landslide remembers those threads' expected {\tt xbegin} results
and refuses to test any others
(unless DPOR separately found them, too, in conflict),
thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.},
available with the {\tt -R} command-line option.

@@

Finally, the keen-eyed reader will notice the state space sizes reported here
differ from those reported in \cite{sigbovik-htm}.
All experiments have been re-run
on account of three updates to Landslide's exploration algorithm
implemented since then:
the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commit 0447666),
the {\tt thrlib\_function} directive
to mark internal thread library logic as trusted (\sect{\ref{sec:landslide-config-landslide}}, commit 64a02e4),
and fixing a soundness bug in which Landslide could neglect to inject transaction failures
immediately after a thread switch (commit dcae85b).
On account of the former two updates, some state spaces may be smaller than before;
on account of the third, some may be larger.
These updates do not discredit the bug-finding results (a bug is a bug),
but the previously-published verification results should be considered outdated.

@@

			& 4,2 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{todo atm}	& \cpu{3.28}	& \ints{1} \\

@@

In the case of {\tt htm2},
it is easy to look at the program with human intuition and judge that,
because the protocol's only state is stored in a single boolean,
with no use of unbounded-capacity data structures or 
contention-dependent exponential backoff loops,
it would be unimaginable that adding a 5th thread to the system
could make any difference in correctness where 4 threads could not,
or that a 4th repetition between 2 threads could make a difference where 3 could not,
and ultimately that it must safely generalize to all $(K, N)$.
% under certain invisible assumptions like the system runs with SC memory! (spoilers!!)
For more complicated algorithms,
which may involve complex conflict patterns such as tree rebalancing or map resizing,
generalizing the verification is not so straightforward;
other formal verification approaches aside,
the user must ultimately be content with the probabilistic assurance that
as $K$ and $N$ increase,
the likelihood that a bug exists which requires more threads or iterations to expose grows ever lower
% (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
(none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
% and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
% but it's not clear how to state this "analogy" concisely in text

@@

@@

Visual inspection of the AVL tree and separate-chaining map implementations \cite{tm-benchmark-cmu},
after correcting the former's atomicity bug (Figure~\ref{fig:avlbug}),
reveals that every use of HTM followed exactly the same pattern,
running identical data structure logic in both the transactional and abort paths,
as though HTM were merely a mutual exclusion lock with fancy performance characteristics.
% this citation is a lil problematic
Prior work \cite{dbug-phdthesis} proposed {\em abstraction reduction},
in which the user identifies program components that can be separated by a well-understood API,
then tests each one against the API individually,
effectively turning multiplicative state space size factors into additive ones.

In this case, I split the lock-like HTM use
and the mutually-exclusive data structure code
into separate tests,
{\tt lock}, which checks the use of HTM guarantees mutual exclusion,
and {\tt avl\_mutex}/{\tt map\_mutex},
which replace the open-coded HTM use with an already-trusted P2 mutex.
{\tt lock\_fast},
a bonus test,
checks the transactional lock's performance by asserting that
its internal logic won't trigger conflict aborts
even when the client's accesses are independent.
Figure~\ref{fig:htm-lock} shows the core logic of these tests.
Finally, I parameterized them over how the lock was implemented:
{\tt spinlock} ``real-world'' code from \cite{spinlock-rtm-github},
{\tt spin\_fixed} the same with the performance bug from \sect{\ref{sec:tm-eval-bugs}} fixed,
and {\tt mutex} using Landslide-annotated P2 mutexes (as the AVL and map implementations do).

@@

\begin{table}[h]

@@

		%\begin{tabular}{cc}

@@

			%\end{tabular}
			%&
			%\begin{tabular}{cc||r|r}
			%& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			%\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\

@@

@@

		%\end{tabular}

@@

	\caption{Continuation of Table~\ref{tab:tm-verifs}.
		Demonstrates abstraction reduction \cite{dbug-phdthesis}

@@

		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional.

>>>>

Support for TSX programs in Landslide is implemented in four parts, broadly speaking:
the user interface,
plus one internal part corresponding to each proof above,
plus a bonus optimization for pruning equivalent interleavings under the new concurrency model.

@@

Landslide provides its own ``implementation'' of the TSX interface,
which matches GCC's interface exactly,
located in {\tt 410user/inc/htm.h}
under {\tt pebsim/p2-basecode/}.
The interface is implemented in {\tt 410user/libhtm/htm.c},
perhaps surprisingly, as totally empty functions;
Landslide hooks these functions' addresses during instrumentation
and inserts preemption points, failure injections,
et cetera as necessary whenever the execution encounters them.

@@

Finally, Quicksand supports the following command-line options to enable various sets of HTM features within Landslide.
Each option requires each of the previously-listed options to also be present.
\begin{itemize}
	\item {\tt -X} (for ``tsX'' or ``Xbegin'') enables the basic features:
		preemption points on each {\tt \_xbegin()} and {\tt \_xend()} call,
		failure injection on each of the former
		(always returning {\tt \_XABORT\_RETRY} as the failure code)
		(\sect{\ref{sec:txn-failure}}),
		and treatment of transactional regions during data-race analysis (\sect{\ref{sec:txn-datarace}}).
	\item {\tt -A} (for ``xAbort codes'')
		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
	\item {\tt -S} (for ``Stm'' or ``Suppress retries'')
		disables the {\tt \_XABORT\_RETRY} failure reason,
		which causes Landslide to match STM's execution semantics rather than HTM.
		(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
\end{itemize}

@@

While Figure~\ref{fig:retry-set-example}'s example may seem contrived
(what program's transactions would just give up and do no work if aborted?),
it is easy to imagine a larger transactional data structure,
two insertions into which might operate on disjoint nodes or array indices,
allowing simultaneous transactions to usually succeed,
while the more uncommon abort paths might take the opportunity to assert a full consistency check of all elements,
ultimately resulting in a similar conflict pattern.
Also, the evaluation will later show
(\sect{\ref{sec:tm-eval-retrysets}})
that even programs with fully conflicting success/abort paths
may still exhibit some equivalent thread interleavings of this pattern
after their failure paths are split apart into smaller transitions by data-race preemption points.

To solve this,
I extended Landslide
with {\em retry set reduction},
an experimental feature named after the analogous {\em sleep sets} from prior work
(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
if either or both sides of a memory conflict were part of a transaction,
it records a {\em retry set},
i.e., the pair of {\tt xbegin} results executed by the conflicting threads.
% ah-- not compatible after all!
%in addition to tracking the set of abort codes possible at each preemption point.
Later, when traversing that new branch,
Landslide remembers those threads' expected {\tt xbegin} results
and refuses to test any others
(unless DPOR separately found them, too, in conflict),
thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
% TODO: make a real commit hash instead; put on master
The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.},
available with the {\tt -R} command-line option.

@@

Finally, the keen-eyed reader will notice the state space sizes reported here
differ from those reported in \cite{sigbovik-htm}.
All experiments have been re-run
on account of three updates to Landslide's exploration algorithm
implemented since then:
the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commit 0447666),
the {\tt thrlib\_function} and {\tt TRUSTED\_THR\_JOIN} directives
to mark internal thread library logic as trusted (\sect{\ref{sec:landslide-config-landslide}}, commit 64a02e4),
and fixing a soundness bug in which Landslide could neglect to inject transaction failures
immediately after a thread switch (commit dcae85b).
On account of the former two updates, some state spaces may be smaller than before;
on account of the third, some may be larger.
These updates do not discredit the bug-finding results (a bug is a bug),
but the previously-published verification results should be considered outdated.

@@

			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{qqq}	& \ETAdag{todo atm}	& \cpu{3.28}	& \ints{1} \\

@@

In the case of {\tt htm2},
it is easy to look at the program with human intuition and judge that,
because the protocol's only state is stored in a single boolean,
with no unbounded-capacity data structures or
contention-dependent exponential backoff loops,
it would be unimaginable that adding a 5th thread to the system
could make any difference in correctness where 4 threads could not,
or that a 4th repetition between 2 threads could make a difference where 3 could not,
and ultimately that it must safely generalize to all $(K, N)$.
% under certain invisible assumptions like the system runs with SC memory! (spoilers!!)
Generalizing the verification is not so straightforward
for more complicated algorithms,
which may involve complex conflict patterns such as tree rebalancing or map resizing;
other formal verification approaches aside,
the user must ultimately be content with the probabilistic assurance that
as verified $K$ and $N$ increase,
the likelihood that a bug exists which requires more threads or iterations to expose grows ever lower
% (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
% and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
% but it's not clear how to state this "analogy" concisely in text

@@

\subsubsection{Abort code (STM) reduction}
\label{sec:tm-eval-stm}

@@

Visual inspection of the AVL tree and separate-chaining map implementations \cite{tm-benchmark-cmu},
after correcting the former's atomicity bug (Figure~\ref{fig:avlbug}),
reveals that every use of HTM followed exactly the same pattern:
running identical data structure logic in both the transactional and abort paths,
as though HTM were merely a mutual exclusion lock with fancy performance characteristics.
% this citation is a lil problematic
Prior work \cite{dbug-phdthesis} proposed {\em abstraction reduction},
in which the user identifies program components that can be separated by a well-understood API,
then tests each one against the API individually,
effectively turning multiplicative state space size factors into additive ones.

In this case, I split the lock-like HTM use
and the mutually-exclusive data structure code
into separate tests,
{\tt lock}, which checks that the use of HTM guarantees mutual exclusion,
and {\tt avl\_mutex}/{\tt map\_mutex},
which replace the open-coded HTM use with an already-trusted P2 mutex.
{\tt lock\_fast},
a bonus test,
checks the transactional lock's performance by asserting that
its internal logic won't trigger conflict aborts
even when the client's accesses are independent.
Figure~\ref{fig:htm-lock} shows their core logic.
Finally, I parameterized them over how the lock was implemented:
a real-world {\tt spinlock} implementation from \cite{spinlock-rtm-github},
{\tt spin\_fixed}, the same with the performance bug from \sect{\ref{sec:tm-eval-bugs}} fixed,
and {\tt mutex}, using Landslide-annotated P2 mutexes (as the AVL and map implementations do).

@@

\begin{table}[t]

@@

		\begin{tabular}{cc}

@@

			&

@@

			\\
			\\

@@

		\\
		\\
		(a) Verifying HTM locks alone.
		&
		(b) Verifying the lock's client code.

@@

	\caption{Continuation of Table~\ref{tab:tm-verifs},
		demonstrating abstraction reduction \cite{dbug-phdthesis}

@@

		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional at all.
